Automatic extraction of symbolic network signatures from malware code

### 2016-11-21 paper.tex
Progress report

\section{Project Proposal}
\textbf{Tasks}:
\begin{itemize}
	\item Identify and characterize features of common malware code - 1-2 weeks
	\added{Refer to ---}
	\item \deleted{Develop the abstract semantics and operators for extracting  
	symbolic network signatures from malware code - 2-3 weeks} \\
	\added{Automatically generate a symbolic automaton of the given program
		based on packet and variable state - 2-3 weeks}
	\item Modifying/Extending existing tools like Codesurfer for symbolic 
	network signature extraction - 2-3 weeks
\end{itemize}

@@

\appendix
\section{Malware program model}
The malware program listens to incoming packets, and the
network signature is the set of incoming packets which triggers
an attack by the program. We propose a reasonable abstraction
of the code of malware programs.

We model a malware program as a infinite reactive system 
of the form:
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
}
\end{lstlisting}
Each iteration of the loop, the program receives a
packet, and based on current state and packet received,
it performs certain actions. Using static analysis
techniques, we can identify two set of variables, \emph{state}
and \emph{packet} variables. State variables represent the 
program state in terms of the protocol, which packet variables
are used to represent the current packet being processed by
the program.
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
	if (a == 1 and packet[8] = "b") {
		a = a + 1;
	}
	...
}
\end{lstlisting}	
In the above code snippet, $a$ is a state variable
and $packet$ is a packet variable. This also
illustrates the packet processing 
function of the program, which based on the current
input packet and state variables, either transitions
to a new program state or performs an action (attack or
send packets). 
\begin{lstlisting}	
a = 0;
while(True) {
packet = socket.receive();
	if (a == 0 and packet[0] = "a") {
		a = a + 1;
	}
	if (a == 1 and packet[0] = "b") {
		a = a + 1;
	}
	if (a == 2 and packet[0] = "c") {
		attack()  [the attack point is provided as input to analysis]
	}	
}
\end{lstlisting}	
The above code snippet represents the complete model
of the malware program, where based on input packets and
current state, the program transitions to a new state. 
The symbolic automaton for this program is 
shown in \Cref{fig:progautomaton}. Network signatures
for the malware are all paths from start state ($a=0$)
to state of attack. In this case, the sequence of packets
in order to initiate an attack is $a* \rightarrow b* \rightarrow c*$.
\begin{figure}[h]
	\centering
	\begin{tikzpicture}[shorten >=0.5pt,node distance=3cm,on grid,auto] 
	\node[state, initial] (v0)   {$a = 0$}; 
	\node[state] (v1) [right=of v0] {$a = 1$}; 
	\node[state] (v2) [right=of v1] {$a = 2$};
	\node[state] (v3) [right=of v2] {$attack$}; 	
	\path[->] 
	(v0) edge node {$a*$} (v1)
	(v1) edge node {$b*$} (v2)
	(v2) edge node {$c*$} (v3);
	\end{tikzpicture}
	\caption{Symbolic automaton for malware program.}
	\label{fig:progautomaton}
\end{figure}
\bibliographystyle{abbrv}
\bibliography{references}

>>>>

\section{Project Proposal}
\textbf{Tasks}:
\begin{itemize}
	\item Identify and characterize features of common malware code - 1-2 weeks.
	Refer to \Cref{sec:malwaremodel}
	\item \added{Extracting a network signature of length $n$ 
		from the program \\
	Refer to \Cref{sec:extraction}
	\item  Generalizing the network signature extraction and further optimizations
	and evaluation.}
	\item \deleted{Modifying/Extending existing tools like Codesurfer for symbolic 
		network signature extraction - 2-3 weeks}
\end{itemize}

@@

\appendix
\section{Malware program model} \label{sec:malwaremodel}
The malware program listens to incoming packets, and the
network signature is the set of incoming packets which triggers
an attack by the program. We propose a reasonable abstraction
of the code of malware programs.

We model a malware program as a infinite reactive system 
of the form:
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
}
\end{lstlisting}
Each iteration of the loop, the program receives a
packet, and based on current state and packet received,
it performs certain actions. Using static analysis
techniques, we can identify two set of variables, \emph{state}
and \emph{packet} variables. State variables represent the 
program state in terms of the protocol, which packet variables
are used to represent the current packet being processed by
the program.
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
	if (a == 1 and packet[8] = "b") {
		a = a + 1;
	}
	...
}
\end{lstlisting}	
In the above code snippet, $a$ is a state variable
and $packet$ is a packet variable. This also
illustrates the packet processing 
function of the program, which based on the current
input packet and state variables, either transitions
to a new program state or performs an action (attack or
send packets). 
\begin{lstlisting}	
a = 0;
while(True) {
packet = socket.receive();
	if (a == 0 and packet[0] == "a") {
		a = a + 1;
	}
	if (a == 1 and packet[0] == "b") {
		a = a + 1;
	}
	if (a == 2 and packet[0] == "c") {
		attack()  [the attack point is provided as input to analysis]
	}	
}
\end{lstlisting}	
The above code snippet represents the complete model
of the malware program, where based on input packets and
current state, the program transitions to a new state. 
The symbolic automaton for this program is 
shown in \Cref{fig:progautomaton}. Network signatures
for the malware are all paths from start state ($a=0$)
to state of attack. In this case, the sequence of packets
in order to initiate an attack is $a* \rightarrow b* \rightarrow c*$.
\begin{figure}[h]
	\centering
	\begin{tikzpicture}[shorten >=0.5pt,node distance=3cm,on grid,auto] 
	\node[state, initial] (v0)   {$a = 0$}; 
	\node[state] (v1) [right=of v0] {$a = 1$}; 
	\node[state] (v2) [right=of v1] {$a = 2$};
	\node[state] (v3) [right=of v2] {$attack$}; 	
	\path[->] 
	(v0) edge node {$a*$} (v1)
	(v1) edge node {$b*$} (v2)
	(v2) edge node {$c*$} (v3);
	\end{tikzpicture}
	\caption{Symbolic automaton for malware program.}
	\label{fig:progautomaton}
\end{figure}

\section{Signature Extraction} \label{sec:extraction}
We now describe a procedure to generate a network signature
of length $n$ (length of the signature is the number of packets
received to trigger attack) for a boolean malware program. 
We use the SMT solver Z3~\cite{z3}
to find valid network signatures. In this section, we describe
the constraints generated for the SMT solver from the program
such that the model generated by the solver can be used to 
extract signatures.

\subsection{Preliminaries}
For a malware program $M$, let $S$ denote
 the vector of boolean state variables in the program,
  and $P$ denote the vector of boolean packet variables
  in the program\footnote{
  	For example, a 60 bit packet is a vector 
  	of 60 boolean variables.}. Let the
  initial values of the state variables be $S_{0}$.

As described in the earlier section, the malware
packet processing code is a set of \texttt{if} 
statements whose predicates are over the set of state
and packet variables. Let the set of \texttt{if} 
statements be denoted by $I$, and let us denote each \texttt{if} 
statement by $(\gamma, \delta) \in I$, where $\gamma(S,P)$
denotes the predicate on $S$ and $P$, and $\delta(S,P)$ denotes
the action on $S$ if $\gamma(S,P)$ is satisfied. Therefore,
the semantics of the \texttt{if} statement is as follows:
\begin{equation}
\gamma(S,P) \Rightarrow S' = \delta(S,P)
\end{equation}
where $S'$ denotes the new state after the \texttt{if} 
block is executed.

\subsection{Transition Model}
We denote the state
vector after receiving $i$ packets as $S_i$.
When packet $i+1$ of the signature is received, 
mlaware $M$, based on $S_i$ and $P_{i+1}$, will execute 
a \texttt{if} $(\gamma_j, \delta_j)$ block\footnote{
	We assume one \texttt{if} block is executed for every packet.}.
Therefore, we can model the malware state transition 
by the following constraints: 
\begin{equation}
	\forall i. 0 \leq i \leq n-2.~\forall (\gamma, \delta) \in I. ~\gamma(S_i, P_{i+1}) \Rightarrow S_{i+1} = \delta(S_i, P_{i+1})
\end{equation}
Let the attack point block be $(\gamma_M, \delta_M)$. 
Therefore, a network signature of length $n$ means
that $P_n$ triggers the attack. The state of the program
is $S_{n-1}$. The following constraint holds:
\begin{equation}
	\gamma_M(S_{n-1},P_{n})
\end{equation}
The SMT solver will return a model for packets $P_1,
\ldots P_n$ such that the program after these receiving
these packets reaches the attack point. Therefore, 
we run this procedure for varying lengths $n$ to find
network signatures of different lengths.

Also, this procedure only returns a particular signature
of length $n$. We can generalize it by analyzing the 
transitions of $M$ to generalize the signature. For
example, if the $i^{th}$ packet only matches certain 
predicates in the transition, we can generalize the packet on
these bits (the rest of the bits are don't care). To 
generate the different singatures of length $n$, 
we can run the SMT solver multiple times to get different 
models satisfying the constraints.

\subsection{Example}
Let us consider the following program receiving two-bit packets:
\begin{lstlisting}	
s1 = 0;
s2 = 0;
while(True) {
	packet = socket.receive();
	if (s1 == 0 and packet[0] == 0) {
	s1 = 1;
	}
	if (s1 == 1 and s2 == 0 and packet[1] == 1) {
	s2 = (not s2) and packet[1];
	}
	if (s1 == 1 and s2 == 1 and packet[0] == 1 and packet[1] == 1) {
	attack()  
	}	
}
\end{lstlisting}	
We denote the state variables by $s_1, s_2$ and packet 
variables by $p_1, p_2$.
For the first \texttt{if} block, 
$\gamma: \neg s_1 \wedge \neg p_1$, and $\delta: s_1'$ ($s_1$ 
is set by the action). Similarily,
for the second block: 
$\gamma: s_1 \wedge \neg s_2 \wedge p_2$, and $\delta: 
s_2' \Leftrightarrow \neg s_2 \wedge p_1$.
For this program, initial state $S_0 = (0, 0)$.
The attack point predicate $\gamma_M$ is as follows:
\begin{equation}
	s_1 \wedge s_2 \wedge p_1 \wedge p_1
\end{equation}
Therefore, if we add the constraints as described in the 
earlier section, for $n=3$, we would obtain a
valid network signature. 
\bibliography{references}
